### 2013-05-16 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding breakdown analysis to spn chapter

\subsubsection{MCSA Breakdown Analysis}
\label{subsubsec:breakdown_analysis}
For each solver and estimator combination, the spectral radius of the
iteration matrix was varied by changing the absorption cross section
while fixing the grid size and scattering cross section. At each
solve, a minimum of one stochastic history per degree of freedom (DOF)
in the problem was used to compute the Monte Carlo correction. If the
solver could not converge in less than 100 iterations, the number of
histories was increased by increments of 5,000 until convergence was
achieved in less than 100 iterations. The number of iterations
required to converge and the time to converge was recorded.

>>>>

\subsubsection{MCSA Breakdown Analysis}
\label{subsubsec:breakdown_analysis}
For each solver and estimator combination, the spectral radius of the
iteration matrix was varied by changing the absorption cross section
while fixing the grid size and scattering cross section. At each
solve, a minimum of one stochastic history per degree of freedom (DOF)
in the problem was used to compute the Monte Carlo correction. If the
solver could not converge in less than 100 iterations, the number of
histories was increased by increments of 5,000 until convergence was
achieved in less than 100 iterations. The number of iterations
required to converge MCSA and the time to converge was recorded as a
means to capture the breakdown.

Figure~\ref{fig:breakdown_iterations} gives the number of iterations
required to converge for the chosen number of histories per iteration
given by Figure~\ref{fig:breakdown_histories} using the adjoint solver
with the collision and expected value estimators and the forward
solver with the collision estimator. For spectral radii less than
0.97, all MCSA problems converged with 1 history per DOF (10,000 for
this problem) with the number of iterations required to converge
increasing. Near a spectral radius of 0.97, the number of histories
required to converge MCSA in less than 100 iterations takes a dramatic
rise that exhibits neither exponential nor power law behavior. As the
spectral radius approaches 1, the number of histories required becomes
significant and effectively intractable. Even with this simple
diffusion problem, the behavior is consistent with that oberseved for
the fuel assembly problem with $SP_1$ discretization. In that case, we
estimated a spectral radius of $\approx 0.998$, larger than any of the
spectral radii that could be computed within even 90 minutes of
compute time for this simple two dimensional problem. For that
problem, even 1,000,000 histories ($\approx 50$ per DOF) was not
enough to provide convergence.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in]{chapters/spn_equations/breakdown_iterations.pdf}
  \end{center}
  \caption{\textbf{Iterations required to converge as a function of
      spectral radius for the neutron diffusion problem.} \textit{The
      number of histories was increased to achieve convergence in less
      than 100 iterations. At least 10,000 histories were used for
      each calculation.}}
  \label{fig:breakdown_iterations}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in]{chapters/spn_equations/breakdown_histories.pdf}
  \end{center}
  \caption{\textbf{Histories required to converge as a function of
      spectral radius for the neutron diffusion problem.} \textit{The
      number of histories was increased to achieve convergence in less
      than 100 iterations. At least 10,000 histories were used for
      each calculation.}}
  \label{fig:breakdown_histories}
\end{figure}
In addition to the significantly larger number of histories required
for to achieve convergence for ill-conditioned problems another
penalty is paid due to histories that take longer to
compute. Figure~\ref{fig:breakdown_time} gives the CPU time in
seconds required to compute a single random walk averaged over the
entire set of histories run in the calculation over all iterations. As
the spectral radius increases (correlating to a higher ratio of
scattering in the system) the random walk lengths increase, using more
CPU time to finish the computation. Compared to spectral radii of 0.5,
larger spectral radii over 0.97 have histories that require two orders
of magnitude more computation time. This signficant increase in
computation time per history coupled with the significant increase in
the number of histories required to converge is evidence that for
problems with spectral radii above $\approx 0.97$, using MCSA to solve
any problems of interest is completely intractable. We therefore
require a more expansive set of preconditioning techniques to reduce
the eigenvalue spectrum of the $SP_N$ problem into a regime in which
MCSA is more applicable.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=5in]{chapters/spn_equations/breakdown_time.pdf}
  \end{center}
  \caption{\textbf{CPU time per history as a function of spectral
      radius for the neutron diffusion problem.} \textit{As the
      spectral radius grows, so do the length of the random
      walks. Longer random walks require more CPU time to compute.}}
  \label{fig:breakdown_time}
\end{figure}

\clearpage
